Type-Safe Client-Server Communication in Web Applications

### 2014-02-26 paper.tex
Explain choice of Haste and benefits of compiling from STG.

\section{A seamless programming model}
\section{Implementation}
\section{The Haste compiler}
\section{Discussion and related work}

@@

\paragraph{Reactive client-server toolkit} While we believe that our work
represents a step forward in making web applications more robust and easier to
develop, the basic model for updating the application's user-visible state is
still essentially imperative. Implementing an FRP-style\ \cite{ppfrp} GUI
programming model on top of our client-server abstraction may prove a fruitful
way to further improve the programmability of web-based distributed
applications.

@@

\bibitem[Bracker, Gill (2014)]{sunroof}
J. Bracker and A. Gill, Sunroof: A Monadic DSL for Generating JavaScript, Practical Aspects of Declarative Languages, 65-80

\bibitem[Conductance application server (2014)]{conductance}
THE INTERNETS, http://conductance.io

\bibitem[Duetto compiler (2014)]{duetto}
THE INTERNETS, http://leaningtech.com/duetto/

\bibitem[Elliott (2009)]{ppfrp}
C. M. Elliott, Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell

\bibitem[Hedin et al (2014)]{jsflow}
D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld, JSFlow: Tracking Information Flow in JavaScript and its APIs

>>>>

\section{A seamless programming model}
\section{Implementation}
\section{The Haste compiler}

In order to allow the same language to be used on both client and server, we
obviously need some way to compile that language into Javascript. To this end,
we make use of the Haste compiler\ \cite{haste}, developed as a MSc thesis and
extended as part of this work. Haste builds on the GHC compiler to provide
the full Haskell language, including most GHC-specific extensions, in the
browser.

\paragraph{Choosing a compiler} Haste is by no means the only
Javascript-targeting compiler for a purely functional language. In particular,
the GHC-based GHCJS\ \cite{ghcjs} and UHC\ \cite{uhc} compilers are both
capable of compiling standard Haskell into Javascript; the Fay\ \cite{fay}
language was designed from the ground up to target the web space using a subset
of Haskell; and there exist solutions for compiling Erlang\ \cite{jserlang} and
Clean\ \cite{jsclean} to Javascript as well.

In the face of this multitude of options, why did we settle on Haste? We make
the following demands of a candidate source language:

\begin{itemize}
  \item The language must provide a static type system, since one of our
        primary concerns is to reduce defect rates through static typing of
        the client-server communication channel.
  \item The language must be compilable to both Javascript and a format
        suitable for server-side execution as we want our web applications
        to be written as a single, monolithic program.
  \item We also want our language to be practical, in the sense that its output
        should not be disproportionately large or slow. This is more important
        for the client part, where we have to contend with slow connections and
        the overhead of running within an interpreter, than for the server part.
\end{itemize}

Our first criterion rules out the dynamically typed Erlang. The second rules
out Fay which unfortunately lacks critical Haskell features, most notably type
classes.

This still leaves us with four options: Clean, GHCJS, Haste and UHC. However,
the output from GHCJS is prohibitively large - often in the range of several
megabytes for very simple programs - which rules it out for our use case, in
spite of providing slightly better compatibility with low-level GHC features
than Haste. Similarly, the code produced by the UHC Javascript backend tends to
be relatively slow due to it lacking the more mature optimization capabilities
of GHC, as well as being in the same ballpark as GHCJS for size.\ \cite{haste}

This leaves us choosing between Clean and Haste for our client compiler. Since
most of our abstractions, both the cooperative multitasking and the type-safe
client-server communication itself, are easily expressed as monads, well
supported in Haskell thanks to do-notation, the choice fell to Haste.

\paragraph{Implementation} As previously mentioned, Haste offloads much of the
heavy lifting of compilation - parsing, type checking, intermediate code
generation and many optimizations - onto GHC, and takes over code generation
after the STG generation step, at the very end of the compilation process.
STG is the last intermediate representation used by the GHC compiler before
the final code generation takes place and has several benefits:

\begin{itemize}
  \item STG is still a functional intermediate representation, based on the
        lambda calculus. When generating code for a high level target language
        such as Javascript, where functions are first class objects, this
        allows for a higher level translation than when doing traditional
        compilation to lower level targets like stack machines or register
        machines. This in turn allows us to make more efficient use of the
        target language's runtime, leading to smaller, faster code.
  \item In contrast to Haskell itself and GHC's intermediate Core language, STG
        represents `thunks`, the construct used by GHC to implement non-strict
        evaluation, as closures which are explicitly created and evaluated.
        Closures are decorated with a wealth of information, such as their set
        of captured varibles, any type information needed for code generation,
        and so on. While extracting this information manually is not very hard,
        having this done for us means we can get away with a simpler
        compilation pipeline.
  \item The language is very small, essentially only comprising lambda
        abstraction and application, plus primitive operations and facilities
        for calling out to other languages. Again, this allows the Haste
        compiler to be a very simple thing indeed.
  \item Any extensions to the Haskell language implemented by GHC will already
        have been translated into this very simple intermediate format,
        allowing us to support basically any extension GHC supports without
        effort.
\end{itemize}

\section{Discussion and related work}

@@

\paragraph{Reactive client-server GUI toolkit} While we believe that our work
represents a step forward in making web applications more robust and easier to
develop, the basic model for updating the application's user-visible state is
still essentially imperative. Implementing an FRP-style\ \cite{ppfrp} GUI
programming model on top of our client-server abstraction may prove a fruitful
way to further improve the programmability of web-based distributed
applications.

@@

\bibitem[Bracker, Gill (2014)]{sunroof}
J. Bracker and A. Gill: Sunroof: A monadic DSL for generating JavaScript, Practical Aspects of Declarative Languages, 65-80, 2014

\bibitem[Conductance application server (2014)]{conductance}
THE INTERNETS: http://conductance.io

\bibitem[Dijkstra et al (2013)]{uhc}
A. Dijkstra, J. Stutterheim, A. Vermeulen, S. D. Swierstra: Building JavaScript applications with Haskell, Implementation and Application of Functional Languages, pp 37-52, 2013

\bibitem[Domoszlai et al (2011)]{jsclean}
L. Domoszlai, E. Bruël, J. M. Jansen: Implementing a non-strict purely functional language in JavaScript, in Acta Universitatis Sapientiae, 2011

\bibitem[Duetto compiler (2014)]{duetto}
THE INTERNETS: http://leaningtech.com/duetto/

\bibitem[Ekblad (2012)]{haste}
A. Ekblad: Towards a declarative web, MSc thesis, University of Gothenburg, 2012

\bibitem[Elliott (2009)]{ppfrp}
C. M. Elliott: Push-pull functional reactive programming, in Proceedings of the 2nd ACM SIGPLAN symposium on Haskell, 2009

\bibitem[The Fay language (2014)]{fay}
THE INTERNETS: http://fay-lang.org

\bibitem[Guthrie (2014)]{jserlang}
G. Guthrie: http://luvv.ie/2014/01/21/your-transpiler-to-javascript-toolbox/, 2014

\bibitem[Hedin et al (2014)]{jsflow}
D. Hedin, A. Birgisson, L. Bello and A. Sabelfeld: JSFlow: Tracking Information Flow in JavaScript and its APIs, 2014

\bibitem[Nazarov (2012)]{ghcjs}
V. Nazarov: GHCJS Haskell to Javascript Compiler, https://github.com/ghcjs/ghcjs
